---
title: "Final Project - Applied Statistics for Engineering and Scientists II"
subtitle: "Group 6"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
fontsize: 11pt
geometry: margin=1in
---

<!-- Setup -->

```{r setup, include=FALSE}

required_packages <- c("tidyverse", "glmnet", "knitr", "car")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(knitr)
library(car)
library(readxl)

seeds <- list(
  split = 240200L,
  cv = 240300L,
  features = 240400L
)

```

<!-- Helper functions -->

```{r helper function}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

render_table <- function(df, caption, align = NULL) {
  if (is.null(align)) {
    align <- paste(rep("c", ncol(df)), collapse = "")
  }

  kbl(
    df,
    caption = caption,
    booktabs = TRUE,
    align = align
  ) |>
    kable_styling(
      latex_options = c("hold_position", "striped"),
      position = "center",
      full_width = FALSE,
      font_size = 10
    ) |>
    row_spec(0, bold = TRUE)
}

plot_histogram <- function(data, title, x_title, y_title) { 
  plot <- ggplot(
    data.frame(Value = data), 
    aes(x = Value)
  ) + 
    geom_histogram(
      bins = 20, 
      fill = "steelblue", 
      color = "white", 
      alpha = 0.8 
    ) + 
    geom_vline(
      xintercept = mean(data, na.rm = TRUE), 
      color = "blue", 
      linetype = "dashed", 
      linewidth = 0.8
    ) + 
    geom_vline(
      xintercept = median(data, na.rm = TRUE), 
      color = "darkgreen", 
      linetype = "dotted", 
      linewidth = 0.8
    ) + 
    labs(
      title = title, 
      x = x_title, 
      y = y_title
    ) + 
    theme_minimal(base_size = 11) + 
    theme(
      plot.title = element_text(
        hjust = 0.5, 
        face = "bold"
      )
    ) 
  
  return(plot) 
}

ols_stepwise_regression <- function(x_train,
                                    y_train,
                                    direction = "both") {
  
  df <- as.data.frame(x_train)
  df$y <- y_train
  
  full_model <- lm(
    y ~ .,
    data = df
  )
  
  best_model <- step(
    full_model,
    direction = direction,
    trace = 0
  )
  
  return (best_model)
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

regression_metrics <- function(actual, predicted) {
  
  actual <- as.numeric(actual)
  predicted <- as.numeric(predicted)
  
  rmse <- sqrt(
    mean((actual - predicted)^2)
  )
  
  mae <- mean(
    abs(actual - predicted)
  )
  
  r2 <- 1 - (
    sum((actual - predicted)^2) /
    sum((actual - mean(actual))^2)
  )
  
  data.frame(
    RMSE = rmse,
    MAE = mae,
    R2 = r2
  )
}

```

# Part I - Concrete Compressive Strength

## Exploratory Data Analysis

```{r read file}
raw_data <- read_excel("Concrete_Data.xls")
old_column_names <- names(raw_data)
new_column_names <- c("Cement",
                      "BlastFurnaceSlag",
                      "FlyAsh",
                      "Water",
                      "Superplasticizer",
                      "CoarseAggregate",
                      "FineAggregate",
                      "Age",
                      "Strength")
names(raw_data) <- new_column_names
colSums(is.na(raw_data))

```

```{r histogram}

for (name in new_column_names){
  print(plot_histogram(raw_data[[name]], 
                       paste("Distribution of", name), 
                       name,
                       "frequency"))
}

```

```{r pairwise-relationship}



```

```{r correlation}



```

## Data Cleaning

```{r data-processing}

x_train
y_train
x_test
y_test

```

```{r transform}

transform_data <- raw_data

transform_data$Cement <- sqrt(transform_data$Cement)
transform_data$BlastFurnaceSlag <- transform_data$BlastFurnaceSlag
transform_data$FlyAsh <- transform_data$FlyAsh 
transform_data$Water <- (transform_data$Water)^2
transform_data$Superplasticizer <- transform_data$Superplasticizer
transform_data$CoarseAggregate <- transform_data$CoarseAggregate
transform_data$FineAggregate <- transform_data$FineAggregate
transform_data$Age <- log(transform_data$Age)
transform_data$Strength <- sqrt(transform_data$Strength)

for (name in new_column_names){
  print(plot_histogram(transform_data[[name]], 
                       paste("Distribution of", name) , 
                       name,
                       "frequency"))
}

```

## Feature Selection

We choose VIF screening and stepwise selection

Reason: 

### VIF Screening

```{r vif}

library(car)

train_data <- data.frame(
  x_train,
  y_train = y_train
)

vif_model <- lm(
  y_train ~ .,
  data = train_data
)

vif_values <- car::vif(vif_model)

vif_table <- data.frame(
  Predictor = names(vif_values),
  VIF = as.numeric(vif_values)
)

vif_table <- vif_table[order(-vif_table$VIF), ]

vif_table$VIF <- round(vif_table$VIF, 2)

vif_table$Assessment <- cut(
  vif_table$VIF,
  breaks = c(-Inf, 5, 10, Inf),
  labels = c("Acceptable", "High", "Very High")
)

render_table(
  vif_table,
  caption = "Variance Inflation Factor (VIF) of Training Predictors"
)

```
Compare to correlation which was analyzed before.

Cement, BlastFurnaceSlag, FlyAsh, Water, FineAggregate: VIF high -> Emphasize demand for regularization

### Stepwise selection

```{r stepwise}

stepwise_model <- ols_stepwise_regression(
  x_train = x_train,
  y_train = y_train,
  direction = "both"
)

all_predictors <- colnames(x_train)

selected_predictors <- names(coef(stepwise_model))[-1]

selection_table <- data.frame(
  Predictor = all_predictors,
  Selected = ifelse(
    all_predictors %in% selected_predictors,
    "Yes",
    "No"
  )
)

render_table(
  selection_table,
  caption = "Feature Selection Results from Stepwise Regression"
)

```

## Modeling with Regularization

Baseline model: Linear Regression

Regularized models: Ridge, Lasso, Elastic Net

### Linear Regression

```{r linear}

# Prepare training data
train_data <- data.frame(
  x_train,
  y = y_train
)

# Fit baseline OLS model
ols_model <- lm(
  y ~ .,
  data = train_data
)

# Extract coefficients
ols_coef <- data.frame(
  Predictor = names(coef(ols_model)),
  Coefficient = as.numeric(coef(ols_model))
)

ols_coef$Coefficient <- round(
  ols_coef$Coefficient,
  4
)

# Display coefficients
render_table(
  ols_coef,
  caption = "OLS Baseline Model Coefficients"
)

```

### Ride regression

```{r ridge}

ridge_cv <- fit_cv_glmnet(
  x = as.matrix(x_train),
  y = y_train,
  alpha = 0,
  foldid = foldid
)

ridge_lambda <- ridge_cv$lambda.min

ridge_lambda

ridge_coef <- data.frame(
  Predictor = rownames(as.matrix(
    coef(ridge_cv, s = "lambda.min")
  )),
  Coefficient = as.numeric(
    coef(ridge_cv, s = "lambda.min")
  )
)

ridge_coef$Coefficient <- round(
  ridge_coef$Coefficient,
  4
)

render_table(
  ridge_coef,
  caption = "Ridge Regression Coefficients"
)

```

### Lasso regression

```{r lasso}

lasso_cv <- fit_cv_glmnet(
  x = as.matrix(x_train),
  y = y_train,
  alpha = 1,
  foldid = foldid
)

lasso_lambda <- lasso_cv$lambda.min

lasso_lambda

lasso_coef <- data.frame(
  Predictor = rownames(as.matrix(
    coef(lasso_cv, s = "lambda.min")
  )),
  Coefficient = as.numeric(
    coef(lasso_cv, s = "lambda.min")
  )
)

lasso_coef$Coefficient <- round(
  lasso_coef$Coefficient,
  4
)

render_table(
  lasso_coef,
  caption = "Lasso Regression Coefficients"
)

```

### Elastic Net regression

```{r elastic-net}

alpha_grid <- seq(
  0.1,
  0.9,
  by = 0.1
)

elastic_results <- lapply(
  alpha_grid,
  function(alpha) {
    
    cv_fit <- fit_cv_glmnet(
      x = as.matrix(x_train),
      y = y_train,
      alpha = alpha,
      foldid = foldid
    )
    
    data.frame(
      alpha = alpha,
      lambda = cv_fit$lambda.min,
      cv_mse = min(cv_fit$cvm)
    )
  }
)

elastic_results <- do.call(
  rbind,
  elastic_results
)

best_elastic <- elastic_results[
  which.min(elastic_results$cv_mse),
]

best_elastic

elastic_cv <- fit_cv_glmnet(
  x = as.matrix(x_train),
  y = y_train,
  alpha = best_elastic$alpha,
  foldid = foldid
)

elastic_coef <- data.frame(
  Predictor = rownames(as.matrix(
    coef(elastic_cv, s = "lambda.min")
  )),
  Coefficient = as.numeric(
    coef(elastic_cv, s = "lambda.min")
  )
)

elastic_coef$Coefficient <- round(
  elastic_coef$Coefficient,
  4
)

render_table(
  elastic_coef,
  caption = "Elastic Net Regression Coefficients"
)

```

## Evaluation

Metrics: RMSE / MAE / $R^2$

Reason:

```{r evaluation}

ols_pred <- predict(
  ols_model,
  newdata = data.frame(x_test)
)

ols_metrics <- regression_metrics(
  actual = y_test,
  predicted = ols_pred
)

ridge_pred <- predict(
  ridge_cv,
  newx = as.matrix(x_test),
  s = "lambda.min"
)

ridge_metrics <- regression_metrics(
  actual = y_test,
  predicted = ridge_pred
)

lasso_pred <- predict(
  lasso_cv,
  newx = as.matrix(x_test),
  s = "lambda.min"
)

lasso_metrics <- regression_metrics(
  y_test,
  lasso_pred
)

elastic_pred <- predict(
  elastic_cv,
  newx = as.matrix(x_test),
  s = "lambda.min"
)

elastic_metrics <- regression_metrics(
  actual = y_test,
  predicted = elastic_pred
)

```

Selected model:

Reason:

# Part II - ToothGrowth

```{r data anova}

data(ToothGrowth)

ToothGrowth$supp <- factor(ToothGrowth$supp)
ToothGrowth$dose <- factor(ToothGrowth$dose)

model <- lm(len ~ supp * dose, data = ToothGrowth)
model_trans <- lm(log(len) ~ supp * dose, data = ToothGrowth)

qqnorm(residuals(model))
qqline(residuals(model))

qqnorm(residuals(model_trans))
qqline(residuals(model_trans))



plot(fitted(model_trans), residuals(model_trans),
     xlab = "Fitted values",
     ylab = "Residuals")

abline(h = 0, lty = 2)

plot(residuals(model_trans),
     type = "p",
     xlab = "Observation order",
     ylab = "Residual")

abline(h = 0, lty = 2)


```